RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-04 manuscripts/manuscript.Rmd
merge, edit and reply to @hlapp's edits

<!-- FIXME

ADD STUFF ABOUT:

- Rapid growth in phylogenetics data (+citations)
- Rapid growth of research methods and software that use phylogenetic data. Comparative methods in evolution. Ecological phylogenetics. Biodiversity informatics. (+citations)
- More on data sharing challenges and opportunities a la @Drew_2013
- More on the hoops that users need to jump through if trees, data
and other things all need to come from separate files and need to be
joined on taxon labels.

-->

@@

<!-- What do we need to say, if anything, about the current R formats?  about phylobase?  -->

The R language [@R] has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations [@taskview],
<!-- hmm, taskview citation renders as "'CRAN task view' 2014" intended? -->
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package [@Paradis_2004].

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts,
<!-- such as what? we haven't talked about metadata yet. HL -->
and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
<!-- such as what? should walk through an example here. HL -->
but inherits the limitations of the NEXUS format.
<!-- how can information inherit the limitations of a serialization -->
<!-- format if it's added to data read from that format? HL -->

`RNeXML` opens up a rich world of both currently available and rapidly
growing data sets in the NeXML format to the R user and the large ecosystem
of R packages currently available for handling phylogenetic data.  
<!-- Carl, I assume a period goes here, I inserted one -->
R packages and R users frequently add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. We illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  
`RNeXML` also provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatibility of the file type.

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R
formats
<!-- R packages, not formats, right? HL -->
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

<!-- I still question the wisdom of having multiple names for the same
method. At first this might seem user-friendly, but it will just be confusing
and create a document and source code maintenance nightmare. Why not just pick
read.nexml and be done with it? -->
<!-- For whatever reason, R has native support for aliasing function names;
     so I don't see this creating a maintenance/documentation issue.  If I 
     had to pick, I'd go with read_nexml though; a lot of the R developer 
     community strongly opposes the `.` in function names, since it falsely
     mimics the S3 class structure.  - CB -->

@@

```{r}
library("RNeXML")
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```
<!-- The system.file() command is not pretty. Can we not just read in -->
<!-- the file directly? HL -->
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. <!-- This seems redundant to me to what has already been -->
<!-- said. HL
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and metadata, analogous to the `ape::phylo` object from the 
popular `ape` R package [@Paradis_2004]. --> Although new custom R functions and packages 
could be written for using `nexml` objects, for most users it will be
more convienent to use methods of the many already existing R packages for 
phylogenetics. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
<!-- having a function's return type being more or less unpredictable -->
<!-- sounds like a bad design pattern. Is there at least precedence for -->
<!-- this practice in other R packages? HL --> 
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) <!-- first mention -->
<!-- of TreeBase, include URL? or introduce it in intro? --> through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```
<!-- the example makes me wonder how nexml_read() distinguishes -->
<!-- between a URL and a file name as its argument. HL -->
Reading phylogenies from TreeBASE into R was already possible using
the R package `treebase` [@Boettiger_2012]. However, because it relied
on the NEXUS parser provided in the `ape` package, and because of
incompatible flavors of the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Owing to the syntactically strict
definition of the
NeXML standard, the RNeXML package can successfully read every valid
NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
<!-- it can't read? the sentence as is suggests so, and so is a little -->
<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
<!-- can't read are now invalid NeXML files. Which is probably not -->
<!-- what you mean to suggest. HL -->

@@

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```
<!-- same as above - why can this not be one call? HL -->
Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

@@

<!-- in caper there is this method comparative.data to join data
and trees on taxon labels. it would be nice if this merged data
structure could also be passed to nexml_write (or should that be
write.nexml?) -->
<!-- Great observation.  I think we should add a coercion method
from the caper structure to the nexml S4 structure.  Then nexml_write
can automatically handle this. added as issue now  -->

@@

These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].

<!-- how about add_taxa? -->
<!-- A taxa block is added when trees or characters are added.  Yeah,
     I guess it would make sense to be able to add taxa directly,
     (e.g. a NeXML file containing only RDFa metadata about taxa, 
     independent of characters or trees), but I worry about confusing
     people into thinking they must add taxa blocks separately, and
     about when we should merge or not merge taxa blocks of matching
     species. See https://github.com/ropensci/RNeXML/issues/43 
     Added new issue for this, https://github.com/ropensci/RNeXML/issues/85 -->

Instead of packaging trees and characters at the time of writing a
NeXML file, the document can also be built up step by step using the
functions `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object, or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

This allows us to include multiple trees and character matrices to a
single NeXML file.
<!-- perhaps give a motivating use case for why one might -->
<!-- want to do this? HL -->

@@

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
<!-- Good idea-->

`RNeXML` automatically adds basic metadata by default
<!-- what does "by default" mean here? It doesn't seem like calling -->
<!-- any RNeXML function will add such metadata if it's not there -->
<!-- already? HL -->
such as a time-stamp
of when a tree was produced, and an open data license. The latter by
default is the Creative Commons Zero (CC0) public domain waiver, which
is compatible
with deposition in the Dryad data repository, but can be configured
as a different license if needed.

@@

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:
<!-- I notice that other add_* methods take the nexml object as the -->
<!-- last argument, not the first as seems to be the case here. This -->
<!-- is inconsistent and should be fixed. HL -->
```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

>>>>

@@

The R language [@R] has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations [@taskview],
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package [@Paradis_2004].

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts,
<!-- such as what? we haven't talked about metadata yet. HL -->
and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
<!-- such as what? should walk through an example here. HL -->
but inherits the limitations of the NEXUS format.
<!-- how can information inherit the limitations of a serialization -->
<!-- format if it's added to data read from that format? HL -->
<!-- CB: Sorry, was refering to data _not_ read from NEXUS format,
but rather data created in an R format/data-structure.  Information
that is in that R data structure (e.g. a simmap) is lost when using
write.nexus(simmap) -->

`RNeXML` opens up a rich world of both currently available and rapidly
growing data sets in the NeXML format to the R user and the large ecosystem
of R packages currently available for handling phylogenetic data. 
R packages and R users frequently add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. We illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.

`RNeXML` also provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatibility of the file type.

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in data structures
provided by existing R packages
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

@@

```{r}
library("RNeXML")
nex <- nexml_read("components/trees.xml")
```
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. <!-- This seems redundant to me to what has already been 
said. HL -->
<!-- CB: Tried to revise what follows to emphasize that we have a new data structure / object class here,
that has unique merits -->
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and semantic metadata. As such, this new R object format differs
from existing formats such as `ape::phylo` format [@Paradis2004],
which describes trees alone, or any of the `phylobase` classes of object that contain 
trees and traits but more limited metadata [@phylobase].

Although new custom R functions and packages 
could be written for using `nexml` objects, for most users it will be
more convienent to use methods of the many already existing R packages for 
phylogenetics. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
<!-- having a function's return type being more or less unpredictable -->
<!-- sounds like a bad design pattern. Is there at least precedence for -->
<!-- this practice in other R packages? HL -->

<!-- Yes, sadly it is.  Really this is a class definition issue that multiPhylo isn't a
proper generalization of a phylo object, but rather a different class entirely. 
Users who don't like this function can use the get_trees_list() to always be predictable.  -->

While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) 
 through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```
<!-- the example makes me wonder how nexml_read() distinguishes -->
<!-- between a URL and a file name as its argument. HL -->
<!-- just type `nexml_read` and you'll find out ;-).  We use a simple regex  CB -->

Reading phylogenies from TreeBASE into R was already possible using
the R package `treebase` [@Boettiger_2012]. However, because it relied
on the NEXUS parser provided in the `ape` package, and because of
incompatible flavors of the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Owing to the syntactically strict
definition of the
NeXML standard, the RNeXML package can successfully read every 
NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
<!-- it can't read? the sentence as is suggests so, and so is a little -->
<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
<!-- can't read are now invalid NeXML files. Which is probably not -->
<!-- what you mean to suggest. HL -->
<!-- Correct, there are indeed a few invalid ones (e.g. some are purely
erroneous HTML, see issues on supertreebase repo) which it cannot
parse.  But like you say, no need to confuse the reader.  I've dropped the 
word 'valid'.  -->

@@

```{r }
nex <- nexml_read("components/comp_analysis.xml")
get_characters(nex)
```
<!-- same as above - why can this not be one call? HL -->
<!-- CB: I've dropped the system.file, even though that is sadly the 
standard way for R to load data that comes with a package but is not in CSV, R, or .rda format.

Indeed `get_characters` could call the file path directly, but there is key
conceptual reason for not doing this.  Note that all the examples call nexml_read to get a nex object
before using get_characters, get_trees, etc.  That is because we really want the user to get used
to manipulating the nexml R data structure, rather than using the external XML file directly.

The functions embody this design principle deep down.  We could have written get_characters etc
in a way that it just yanked the characters out of the XML and left everything else, rather than 
converting the whole NeXML file into a serialized R object stored in memory.  But the latter strategy
gives us a native R object that people can extend, write functions for, and subset and query with 
the standard R syntax.  Without this, the package isn't really providing a native format, and users 
and developers are left on their own to either also work directly with the XML or just convert immediately
into a more limited R format like ape::phylo.  
-->

Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

@@

@@

These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].
Instead of packaging trees and characters at the time of writing a
NeXML file, the document can also be built up step by step using the
functions `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object, or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

This allows us to include multiple trees and character matrices to a
single NeXML file, for instance, to represent samples from posterior
probability distributions of inferred or simulated phylogenies or
character states.

@@

`RNeXML` automatically adds basic metadata by default
<!-- what does "by default" mean here? It doesn't seem like calling -->
<!-- any RNeXML function will add such metadata if it's not there -->
<!-- already? HL -->
<!-- Indeed it does.  We add the time stamp, default license, and user ID even though
     it's not "there" in the data.  -->
such as a time-stamp
of when a tree was produced, and an open data license. The latter by
default is the Creative Commons Zero (CC0) public domain waiver, which
is compatible
with deposition in the Dryad data repository, but can be configured
as a different license if needed.

@@

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:
<!-- I notice that other add_* methods take the nexml object as the -->
<!-- last argument, not the first as seems to be the case here. This -->
<!-- is inconsistent and should be fixed. HL -->

<!-- Not so, the nexml object is the second argument here too.  If no existing 
nexml object is given, the add function starts a new nexml object.  If one is
given, it appends it's first argument onto the existing nexml, as described above.

Notice that bird.orders is an ape::phylo object, which we are adding to a new 
nexml object.  (R allows for default arguments, the default second argument is 
new("nexml") -->

```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",
